The Spine of American Legal Practice

### 2016-06-20 article/article.Rmd
Basic cleanup from Google Docs

---
title: The Spine of American Legal Practice
author:
- Kellen Funk
- Lincoln Mullen
output: pdf_document
...

@@

@@

**Part 1.**

@@

&gt;&gt;Courage; persevere in the support of written reason against precedents and vague traditions. . . . Let four or five good heads be united in a commission, to frame in silence the project of a code. It is not so difficult a task. It is only to consult together, and to select. Do so with your best authors as we did with ours, and principally with Pothier's treatise on Obligations, which we simply converted into articles of our code.&lt;&lt;

@@

After David Dudley Field and two other lawyers were appointed to the procedural commission, their reports made clear that they favored the French model but understood political realities would hold them to the English model. From 1847 to 1850, the commissioners made five reports to the legislature knowing they had no power to keep legislators from amending their code or even defeating it altogether. Each time they reminded legislators that "public opinion had issued its mandate in the most imposing form" of a constitutional decree. The constitution, together with the legislative act appointing the commission "gave the commissioners instructions so precise, as to leave them no discretion, if they had desired it, \[and\] promised them therefore in advance, so long as they obeyed those instructions, the concurrence and co-operation of all departments of the government."[^16] Although the theory of codification made it a democratic enterprise, in practice Field---an ardent Jacksonian---insisted the democratic legislature ought to defer to the expertise of the commissioners.

@@

**Part 2.**

Although the amount of legislation that governed American civil practice is impressive---even if uncodified, every state had procedure statutes that were amended and re-enacted nearly every decade---a corpus of procedural legislation would not even count among digital historians as "big data." Our collection of relevant (or possibly relevant) laws comprise 135 statutes from the nineteenth century. It includes separately bound codes of civil procedure as well as codes or regulations appearing session laws and statutory compilations.[^30] It comprises some **\[NUMBER\]** pages, **\[NUMBER\]** words, organized into 98,000 regulations. The corpus is not exhaustive. It includes the initial enactment of every code of civil procedure, as well as a sampling of procedure statutes and re-enacted codes from states reputed to have been legally influential, but it does not include every known statute of American procedural law. While a comprehensive project may be illuminating in its own ways, our specific question of how New York legislation influenced other states permits a more curated corpus.

@@

&gt;&gt;§ 554. The distinction between actions at law and suits in equity, and the forms of all such actions and suits, heretofore existing, are abolished; and, there shall be in this state, hereafter, but one form of civil action, for the enforcement or protection of private rights and the redress or prevention of private wrongs, which shall be denominated a civil action. § 555. In such action, the party complaining is known as the plaintiff, and the adverse party as the defendant.&lt;&lt;

@@

&gt;&gt;§ 1. There shall be in this State but one form of Civil Action, for the enforcement or protection of private rights, and the redress or prevention of private wrongs. §2. In such action the party complaining shall be known as the plaintiff, and the adverse party as the defendant.&lt;&lt;

@@

**Part 3.**

@@

Even later New York codes can be considered a separate family. In 1876 a New York commission produced a new code attempting to consolidate all the statutory amendments subsequent to the 1851 Field Code. David Dudley Field was upset by the changes introduced in this revised code, called the Throop Code. A count by a "friend" of Field's found that only three sentences of the Field Code had carried over word-for-word into the Throop Code. Nevertheless, many of the changes were purely formal, and Field speculated that these changes had been made only "so that \[the Code\] should not appear to be the same thing as before." With respect to Mr. Field or his "friend," we were able to find that the connection was between the Throop Code and the Field Code was somewhat stronger than he thought, and the Throop Code was indeed a descendant, if much revised, from the Field Code.

@@

<!-- ![](media/image04.png){width="6.5in" height="3.7916666666666665in"} -->
California's 1851 code borrowed heavily from two sources. The bulk of the code was taken directly from New York's 1849 or 1850 codes (which are difficult to distinguish exactly from one another.[^42] The remainder of the code was borrowed from California's 1850 code. The 1850 version was in turn borrowed from the New York codes \[**might as well show this, esp. since CA1850 is so impressive**\]. As we would expect, the bulk of the borrowings that come from California instead of directly from New York are at the beginning of the code, where the 1850 code had made amendments specific to the borrowing from California. (Many of the non-matching sections are tables of contents.) The main point is that that the majority of the code was lifted more or less directly from New York, with minor amendments specific to California. California made few to no innovations to the Field Code beyond a rearrangement of its provisions.

@@

Take for instance, provisions regulating witness testimony. At common law, parties and interested witnesses were not permitted to testify in their own causes. Field's Code reversed this rule, expanding witness competency as widely as possible: any person "having organs of sense" could now be admitted as a witness in New York, with only the insane and very young children exempted from the rule. As the Code migrated West, however, legislators added racial exclusions to Field's list. The following cluster documents how California's codifiers engrafted earlier prohibitions from midwestern states into Field's Code. Many other codes then evidence a remarkable uniformity with California's text (which later changed only to add "Mongolian" to the list of races). Iowa's Code had much more minor influence, and Wyoming developed the only truly original section which made clear the link to a racist assumption of infantilization of non-white peoples. \[**Figure: Cluster from Kellen**\]

@@

**Part 4**.

@@

As in postbellum North Carolina, establishing a flow of credit through the remedial system became a leading priority of western lawyers. While the so-called new western history has shed significant light on neglected topics of Indian dispossession and environmental management, it has often done so by leaving out of view matters of political economy, a staple of the old western history. As one work in that older tradition argued, "Debt collection was the central part of law practice for the \[western\] bar and remained a key part of private practice throughout the century."[^62] On that understanding, one western lawyer succinctly summarized the difference between the Code and the common law as "whether a merchant had better try to collect a \$500 note or burn it up." Tiring of all the focus on creditors' remedies, one legislator observed that he "never knew one of these professionals who undertook to write up the beauties of the New York code, . . . that he did not also break out somewhere with 'take for instance the case of an action on a promissory note,' as though the collection of notes was about all there could be any law needed for."[^63]

@@

**Conclusion**

@@

[^10]: Friedman, *A History of American Law*, 302. When the intellectual historian Perry Miller developed a reader surveying *The Legal Mind in America*, codification was its only central theme, as Miller argued it was the one intellectual topic that attracted lawyers away from their practices long enough to debate. Perry Miller, ed., *The Legal Mind in America: From Independence to the Civil War* (Anchor 1962). See also Charles M. Cook, *The American Codification Movement: A Study in Antebellum Legal Reform* (1983); Robert W. Gordon, "The American Codification Movement," 36 *Vanderbilt Law Review* 431 (1983); Maurice Eugen Lang, *Codification in the British Empire and America* (Lawbook Exchange 1924). On common theories of codification that transcended jurisdictional boundaries, see Csaba Varga, *Codification as a Socio-Historical Phenomenon* (Budapest, 2d. ed. 2011 \[1991\]); Roger Berkowitz, *The Gift of Science: Leibniz and the Modern Legal Tradition* (Fordham 2010).

@@

[^22]: On the French procedure code, see C.H. van Rhee, *European Traditions in Civil Procedure* (Antwerp 2005). Bentham, the leading proponent of codification in England, argued that procedure was the one department of the law that ought to remain *un*codified. So long as the law of civil and criminal obligations and the law of property were sufficiently codified, a "natural procedure" arising from judicial discretion and flexibility would be superior to "technical" written rules. Later in his career Bentham produced the "Outlines of a Procedure Code" as a "provisional" remedy, but he insisted that a procedure code on its own could not be "invested with the form of law" without "reference to the codes of law, penal and non-penal, to which it has for its object and purpose the giving execution and effect." Although it spanned nearly 200 pages, Bentham's code favored general moral maxims over precise details, for instance: "On each occasion, have constant regard for all the several ends of justice; that is to say, minimize the sum, or the balance of evil." Jeremy Bentham, Principles of Judicial Procedure with the Outline of a Procedure Code, in John Bowring, ed., *Works of Jeremy Bentham* (2d. ed. 1843 \[1839\]), 1-189, preface and 28, ch. 7 § 1. See Lobban, *The Common Law*, 127-131.

@@

[^38]: This algorithm works by extracting a set number of random tokens from each document, allowing the documents to be represented uniformly and compactly. Then those random tokens are grouped into subsets. If any two documents have a matching subset, then they are considered a candidate pair. \[**CLARIFY**\] This algorithm has several useful properties. It approximates the Jaccard similarity of the two documents. It requires a computation for each document rather than each pair of documents, so the computation time grows linearly not geometrically. And by controlling various parameters, one can set a threshold similarity score above which one is likely to find a genuine match and unlikely to find a spurious match.

>>>>

---
title: The Spine of American Legal Practice
author:
- Kellen Funk
- Lincoln Mullen
output: 
  pdf_document:
    number_sections: true
    fig_caption: true
fontsize: 12pt
documentclass: article
...

@@

![The 1861 Morris bill which adapted California's code for Nevada.](morris-bill.jpg)

@@

# Part 1

@@

> Courage; persevere in the support of written reason against precedents and vague traditions. ... Let four or five good heads be united in a commission, to frame in silence the project of a code. It is not so difficult a task. It is only to consult together, and to select. Do so with your best authors as we did with ours, and principally with Pothier's treatise on Obligations, which we simply converted into articles of our code.

@@

After David Dudley Field and two other lawyers were appointed to the procedural commission, their reports made clear that they favored the French model but understood political realities would hold them to the English model. From 1847 to 1850, the commissioners made five reports to the legislature knowing they had no power to keep legislators from amending their code or even defeating it altogether. Each time they reminded legislators that "public opinion had issued its mandate in the most imposing form" of a constitutional decree. The constitution, together with the legislative act appointing the commission "gave the commissioners instructions so precise, as to leave them no discretion, if they had desired it, [and] promised them therefore in advance, so long as they obeyed those instructions, the concurrence and co-operation of all departments of the government."[^16] Although the theory of codification made it a democratic enterprise, in practice Field---an ardent Jacksonian---insisted the democratic legislature ought to defer to the expertise of the commissioners.

@@

# Part 2

Although the amount of legislation that governed American civil practice is impressive---even if uncodified, every state had procedure statutes that were amended and re-enacted nearly every decade---a corpus of procedural legislation would not even count among digital historians as "big data." Our collection of relevant (or possibly relevant) laws comprise 135 statutes from the nineteenth century. It includes separately bound codes of civil procedure as well as codes or regulations appearing session laws and statutory compilations.[^30] It comprises some TODO pages, TODO words, organized into 98,000 regulations. The corpus is not exhaustive. It includes the initial enactment of every code of civil procedure, as well as a sampling of procedure statutes and re-enacted codes from states reputed to have been legally influential, but it does not include every known statute of American procedural law. While a comprehensive project may be illuminating in its own ways, our specific question of how New York legislation influenced other states permits a more curated corpus.

@@

> § 554. The distinction between actions at law and suits in equity, and the forms of all such actions and suits, heretofore existing, are abolished; and, there shall be in this state, hereafter, but one form of civil action, for the enforcement or protection of private rights and the redress or prevention of private wrongs, which shall be denominated a civil action. § 555. In such action, the party complaining is known as the plaintiff, and the adverse party as the defendant.

@@

> § 1. There shall be in this State but one form of Civil Action, for the enforcement or protection of private rights, and the redress or prevention of private wrongs. §2. In such action the party complaining shall be known as the plaintiff, and the adverse party as the defendant.

@@

# Part 3

@@

Even later New York codes can be considered a separate family. In 1876 a New York commission produced a new code attempting to consolidate all the statutory amendments subsequent to the 1851 Field Code. David Dudley Field was upset by the changes introduced in this revised code, called the Throop Code. A count by a "friend" of Field's found that only three sentences of the Field Code had carried over word-for-word into the Throop Code. Nevertheless, many of the changes were purely formal, and Field speculated that these changes had been made only "so that [the Code] should not appear to be the same thing as before." With respect to Mr. Field or his "friend," we were able to find that the connection was between the Throop Code and the Field Code was somewhat stronger than he thought, and the Throop Code was indeed a descendant, if much revised, from the Field Code.

@@

<!-- ![](media/image04.png){width="6.5in" height="3.7916666666666665in"} -->
California's 1851 code borrowed heavily from two sources. The bulk of the code was taken directly from New York's 1849 or 1850 codes (which are difficult to distinguish exactly from one another.[^42] The remainder of the code was borrowed from California's 1850 code. The 1850 version was in turn borrowed from the New York codes **TODO: might as well show this, esp. since CA1850 is so impressive**. As we would expect, the bulk of the borrowings that come from California instead of directly from New York are at the beginning of the code, where the 1850 code had made amendments specific to the borrowing from California. (Many of the non-matching sections are tables of contents.) The main point is that that the majority of the code was lifted more or less directly from New York, with minor amendments specific to California. California made few to no innovations to the Field Code beyond a rearrangement of its provisions.

@@

Take for instance, provisions regulating witness testimony. At common law, parties and interested witnesses were not permitted to testify in their own causes. Field's Code reversed this rule, expanding witness competency as widely as possible: any person "having organs of sense" could now be admitted as a witness in New York, with only the insane and very young children exempted from the rule. As the Code migrated West, however, legislators added racial exclusions to Field's list. The following cluster documents how California's codifiers engrafted earlier prohibitions from midwestern states into Field's Code. Many other codes then evidence a remarkable uniformity with California's text (which later changed only to add "Mongolian" to the list of races). Iowa's Code had much more minor influence, and Wyoming developed the only truly original section which made clear the link to a racist assumption of infantilization of non-white peoples.

**Figure: Cluster from Kellen**

@@

# Part 4

@@

As in postbellum North Carolina, establishing a flow of credit through the remedial system became a leading priority of western lawyers. While the so-called new western history has shed significant light on neglected topics of Indian dispossession and environmental management, it has often done so by leaving out of view matters of political economy, a staple of the old western history. As one work in that older tradition argued, "Debt collection was the central part of law practice for the [western] bar and remained a key part of private practice throughout the century."[^62] On that understanding, one western lawyer succinctly summarized the difference between the Code and the common law as "whether a merchant had better try to collect a \$500 note or burn it up." Tiring of all the focus on creditors' remedies, one legislator observed that he "never knew one of these professionals who undertook to write up the beauties of the New York code, . . . that he did not also break out somewhere with 'take for instance the case of an action on a promissory note,' as though the collection of notes was about all there could be any law needed for."[^63]

@@

# Conclusion

@@

[^10]: Friedman, *A History of American Law*, 302. When the intellectual historian Perry Miller developed a reader surveying *The Legal Mind in America*, codification was its only central theme, as Miller argued it was the one intellectual topic that attracted lawyers away from their practices long enough to debate. Perry Miller, ed., *The Legal Mind in America: From Independence to the Civil War* (Anchor 1962). See also Charles M. Cook, *The American Codification Movement: A Study in Antebellum Legal Reform* (1983); Robert W. Gordon, "The American Codification Movement," 36 *Vanderbilt Law Review* 431 (1983); Maurice Eugen Lang, *Codification in the British Empire and America* (Lawbook Exchange 1924). On common theories of codification that transcended jurisdictional boundaries, see Csaba Varga, *Codification as a Socio-Historical Phenomenon* (Budapest, 2d. ed. 2011 [1991]); Roger Berkowitz, *The Gift of Science: Leibniz and the Modern Legal Tradition* (Fordham 2010).

@@

[^22]: On the French procedure code, see C.H. van Rhee, *European Traditions in Civil Procedure* (Antwerp 2005). Bentham, the leading proponent of codification in England, argued that procedure was the one department of the law that ought to remain *un*codified. So long as the law of civil and criminal obligations and the law of property were sufficiently codified, a "natural procedure" arising from judicial discretion and flexibility would be superior to "technical" written rules. Later in his career Bentham produced the "Outlines of a Procedure Code" as a "provisional" remedy, but he insisted that a procedure code on its own could not be "invested with the form of law" without "reference to the codes of law, penal and non-penal, to which it has for its object and purpose the giving execution and effect." Although it spanned nearly 200 pages, Bentham's code favored general moral maxims over precise details, for instance: "On each occasion, have constant regard for all the several ends of justice; that is to say, minimize the sum, or the balance of evil." Jeremy Bentham, Principles of Judicial Procedure with the Outline of a Procedure Code, in John Bowring, ed., *Works of Jeremy Bentham* (2d. ed. 1843 [1839]), 1-189, preface and 28, ch. 7 § 1. See Lobban, *The Common Law*, 127-131.

@@

[^38]: This algorithm works by extracting a set number of random tokens from each document, allowing the documents to be represented uniformly and compactly. Then those random tokens are grouped into subsets. If any two documents have a matching subset, then they are considered a candidate pair. **TODO: CLARIFY** This algorithm has several useful properties. It approximates the Jaccard similarity of the two documents. It requires a computation for each document rather than each pair of documents, so the computation time grows linearly not geometrically. And by controlling various parameters, one can set a threshold similarity score above which one is likely to find a genuine match and unlikely to find a spurious match.
